---
title: "Big Data avec R"
author: "S.Bord, T.Mary-Huard"
date: "22 juin 2018"

output:
  pdf_document:
    toc: true
    number_sections: true
    

---


# Load packages
```{r, eval=FALSE, echo=TRUE}
install.packages("nycflights13",  repos='https://cran.univ-paris1.fr/')
install.packages("sparklyr",  repos='https://cran.univ-paris1.fr/')
library(sparklyr)
spark_install(version = "2.2.0")
library(sparklyr)
library(dplyr)
library(pryr) 
```

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
install.packages("nycflights13",  repos='https://cran.univ-paris1.fr/')
install.packages("sparklyr",  repos='https://cran.univ-paris1.fr/')
library(sparklyr)
spark_install(version = "2.2.0")
library(sparklyr)
library(dplyr)
library(pryr) 
```

# Parameters
```{r, eval=TRUE, echo=TRUE}
RepName <- 'C:/Users/sbord/Dropbox/Big_Data/TP'
setwd(RepName)
``` 


# Local space connection
```{r, eval=TRUE, echo=TRUE}
sc <- spark_connect("local", version = "2.2.0")
```


## Add a table in "sc" : copy_to
```{r, eval=TRUE, echo=TRUE}
flights_tbl <- dplyr::copy_to(sc, nycflights13::flights, "flights", overwrite=T) 
dplyr::src_tbls(sc)
ls()
class(flights_tbl)
```
 

## Data manipulation : fligths example

### Selection of fligths 
```{r, eval=TRUE, echo=TRUE}
flight.sel <- flights_tbl %>%  filter(carrier %in% c("B6", "DL", "EV"))
class(flight.sel)
```

### Save the selection in "sc" space
```{r, eval=TRUE, echo=TRUE}
sdf_register(flight.sel, "flightsel")
src_tbls(sc)
```

### Save the selection in "R" environment
```{r, eval=TRUE, echo=TRUE}
flightsel.inR <- flight.sel %>% collect()
```

### Graph: boxplot of delay by carrier
```{r, eval=TRUE, echo=TRUE}
library(ggplot2)
ggplot(flightsel.inR, aes(x=carrier, y=arr_delay)) +  geom_boxplot()
```

### ANOVA of delay by carrier
```{r, eval=TRUE, echo=TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=65)}
resanova <-flight.sel %>% na.omit() %>% ml_linear_regression(x=., response = "arr_delay" , features = "carrier")

summary(resanova)
```



### Principal component analysis (PCA)

#### Selection of variables for PCA
```{r, eval=TRUE, echo=TRUE}
test_tbl <- flights_tbl %>%
  select(one_of(c('dep_time','sched_dep_time','dep_delay',
                  'arr_time','sched_arr_time','arr_delay','carrier')))


DataForPca <- test_tbl %>%  na.omit() %>% filter(carrier=='UA') %>%  select(1:6) 
```

#### Correlation Matrix between PCA variables
```{r, eval=TRUE, echo=TRUE}
CorrMat <- ml_corr(DataForPca, method = "pearson")
class(CorrMat)
CorrMat
```

#### PCA
```{r, eval=TRUE, echo=TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=65)}
PcaResults <- ml_pca(DataForPca)

Coordinates <- sdf_project(PcaResults,DataForPca,features=rownames(PcaResults$pc)) %>%
  select(starts_with('PC')) 

Coordinates %>%
  select(1:2) %>%
  collect() %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(PC1, PC2))  +
  labs(x = "First principal component",
       y = "Second principal component",
       title = "My first (and ugly) PCA with spark"  )
PcaResults
```

#### Kmeans
```{r, eval=TRUE, echo=TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=65)}
KmeansResults <- ml_kmeans(DataForPca, centers=3, iter.max = 100, features=rownames(PcaResults$pc))
arrange(KmeansResults$centers, arr_delay)

Labels <- ml_predict(KmeansResults, DataForPca) %>%
  select(prediction) %>%
  collect()
table(Labels$prediction)


CoordinatesWithLabels <- Coordinates %>% collect() %>% mutate(Labels = Labels$prediction)

CoordinatesWithLabels %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(PC1, PC2, color=Labels))  +
  labs(x = "First principal component",
       y = "Second principal component",
       title = "My second (and ugly) PCA with labels points"  )
```

## Export "sc" data in a file
```{r, eval=TRUE, echo=TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=65)}
filename_to_save <- paste0(RepName, "/export_exemple.csv")
spark_write_csv(DataForPca, path=filename_to_save, header = FALSE, delimiter = ";")
```

# Memory allocation

```{r, eval=TRUE, echo=TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=65)}
sc %>% spark_context %>% invoke("getRDDStorageInfo")
url <- sc %>% spark_context %>% invoke("uiWebUrl") %>% invoke("get")
browseURL(paste(url, "storage", sep="/"))
app_id <-  sc %>% spark_context %>% invoke("applicationId")
httr::GET(paste(url, "api", "v1", "applications", app_id, "storage", "rdd",  sep="/"))
```

# Local space disconnection
```{r, eval=TRUE, echo=TRUE}
spark_disconnect(sc)
```





