---
title: "Interactive dataviz on the web with R & plotly <br> (adapted from @cpsievert's book)"
author: "State of the R - Atelier R de l'AG MIA/NUMM"
venue: "AG MIA"
type: "tutorial"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "main.css"]
    lib_dir: libs
    chakra: "libs/remark-latest.min.js"
    mathjax: null
    nature:
      # approx the golden ratio
      ratio: '14.6:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
date: "2019-05-22"

---
background-image: url(figs/workflow.svg)
background-size: contain
class: inverse

```{r load-packages, echo = FALSE, message = FALSE, warning=FALSE, results='hide', cache=FALSE}
library(tidyverse)
library(plotly)
library(leaflet)
library(DT)
library(crosstalk)
data("txhousing")
data("diamonds")
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  collapse = TRUE,
  fig.width = 12,
  fig.height = 6
)
generate_html <- TRUE
```


<style>
.principles {
  font-size: 150%;
}
</style>

# Data science workflow

---
# Typical example 

.pull-left[
```{r plot-example}
ggplot(iris, 
       aes(x = Sepal.Length, 
           y = Sepal.Width, 
           color = Species)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) +
  theme_minimal()
```
]
.pull-right[
```{r ref.label = 'plot-example', echo = FALSE, fig.retina=3, eval=TRUE}
```
]

---
background-image: url(figs/workflow1.svg)
background-size: contain
class: inverse

# Web graphics are great for communication!

---

A Typically French Example

[Résultats des élections législatives de 2017](https://abonnes.lemonde.fr/data/france/legislatives-2017/)

---
background-image: url(figs/workflow2.svg)
background-size: contain
class: inverse

# Interactive graphics great for exploration!

---
class: middle

# Interactive graph example

.pull-left[
```{r show-diamonds, eval=TRUE}
diamonds %>% count(cut, clarity) %>% 
  DT::datatable(list(
    dom = "tp",
    pageLength = 6))
```
]
.pull-right[
```{r plot-diamonds, echo = FALSE, fig.retina=3, fig.width = 7, eval=TRUE}
plot_ly(diamonds, x = ~cut, color = ~clarity, colors = "Accent") %>% 
  add_histogram()
```
]

---
class: inverse, middle
# How to combine both?

.pull-left[
### Rapid iteration with extensibility & reproducibility

<img src="figs/workflow2.svg" width="100%" >

]

.pull-right[
### Simple, scalable, and secure hosting

<img src="figs/workflow1.svg" width="100%" >
]

---
class: middle, principles, inverse

# A simple solution: **plotly for R**

1. Powered by [plotly.js](https://github.com/plotly/plotly.js), a JavaScript-library for creating interactive graphs. 

2. `plot_ly()`: a direct interface to plotly.js with abstractions to reduce typing (similar to `ggplot2`)

3. `ggplotly()`: translates `ggplot2` to plotly objects. 

---

# Simple example

```{r plotly-start, eval=TRUE}
plot_ly(diamonds, x = ~cut, color = ~clarity, colors = "Accent") %>% 
  add_histogram()
```

---

<img src="figs/printing.svg" width="100%" >

---
class: inverse, center, middle

# Interactive ggplots

---
### Replace plot with ggplotly

```{r diamonds_basic}
library(plotly)
p <- ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_hex(bins = 100)
ggplotly(p)
```

```{r save_diamonds_basic, include=FALSE, eval = generate_html}
p <- ggplot(diamonds, aes(x = log(carat), y = log(price))) + 
  geom_hex(bins = 100)
htmlwidgets::saveWidget(ggplotly(p, width = 800, height = 350), 
                        file = "01_diamonds_basic.html", 
                        selfcontained = F, libdir = "libs")
```

<iframe src="01_diamonds_basic.html" width = "100%" height = "500" seamless="seamless" frameBorder="0"> </iframe>

---
### Works out of the box with ggplot core geom_*

```{r diamonds_density}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + 
    facet_grid(~cut)
ggplotly(p)
```

```{r save_diamonds_density, include=FALSE, eval = generate_html}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + 
    facet_grid(~cut)
htmlwidgets::saveWidget(ggplotly(p, width = 800, height = 350), "02_diamonds_density.html", selfcontained = F, libdir = "libs")
```

<iframe src="02_diamonds_density.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
class: inverse, center, middle

# A few features of plotly: tooltip and highlight

---
### Tooltip

```{r diamonds_tooltip}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + 
    facet_grid(~cut)
*ggplotly(p, tooltip = c("colour", "x")) # color != colour
```

```{r save_diamonds_tooltip, include=FALSE, eval = generate_html}
p <- ggplot(diamonds, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + 
    facet_grid(~cut)
htmlwidgets::saveWidget(ggplotly(p, width = 800, height = 350, tooltip = c("colour", "x")), 
                        "03_diamonds_tooltip.html", 
                        selfcontained = F, libdir = "libs")
```

<iframe src="03_diamonds_tooltip.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Highlight

```{r diamonds_hl}
{{diams <- highlight_key(diamonds, ~clarity)}}
p <- ggplot(diams, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + facet_grid(~cut)
gg <- ggplotly(p, tooltip = c("colour"))
{{highlight(gg, on = "plotly_click")}}
```

```{r save_diamonds_hl, include=FALSE, eval = generate_html}
diams <- highlight_key(diamonds, ~clarity)
p <- ggplot(diams, aes(x = log(price), color = clarity)) + 
    geom_freqpoly(stat = "density") + facet_grid(~cut)
htmlwidgets::saveWidget(highlight(ggplotly(p, width = 800, height = 350), 
                                  on = "plotly_click"), 
                        "04_diamonds_hl.html", selfcontained = F, libdir = "libs")
```

<iframe src="04_diamonds_hl.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### A better use of higlight

```{r texas_housing, eval=TRUE}
DT::datatable(txhousing)
```

---
### A better use of higlight

```{r texas_hl}
tx <- highlight_key(txhousing, ~city)
p <- ggplot(tx) + geom_line(aes(date, median, group = city, text = city))
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350)
highlight(gg, on = "plotly_click")
```

```{r save_texas_hl, include=FALSE, eval = generate_html}
tx <- highlight_key(txhousing, ~city)
p <- ggplot(tx) + geom_line(aes(date, median, group = city, text = city))
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350)
htmlwidgets::saveWidget(highlight(gg, on = "plotly_click"), 
                        "05_texas_hl.html", selfcontained = F, libdir = "libs")
```

<iframe src="05_texas_hl.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Dynamic highlighting 

```{r texas_dynamic_hl}
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350)
highlight(gg, on = "plotly_hover", selectize = TRUE, dynamic = TRUE)
```

```{r save_texas_dynamic_hl, include=FALSE, eval = generate_html}
tx <- highlight_key(txhousing, ~city)
p <- ggplot(tx) + geom_line(aes(date, median, group = city, text = city))
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350)
htmlwidgets::saveWidget(highlight(gg, on = "plotly_hover", selectize = TRUE, dynamic = TRUE), 
                        "06_texas_dynamic_hl.html", 
                        selfcontained = F, libdir = "libs")
```

<iframe src="06_texas_dynamic_hl.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Persistent highlighting 

```{r texas_persistent_hl}
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350)
highlight(gg, on = "plotly_hover", selectize = TRUE, dynamic = TRUE,
          {{persistent = TRUE, defaultValues = "Paris"}}
          )
```

```{r save_texas_persistent_hl, include=FALSE, eval = generate_html}
tx <- highlight_key(txhousing, ~city)
p <- ggplot(tx) + geom_line(aes(date, median, group = city, text = city))
gg <- ggplotly(p, tooltip = "text", width = 800, height = 350) %>% 
  highlight(on = "plotly_hover", selectize = TRUE, 
            persistent = TRUE, defaultValues = "Paris", 
            dynamic = TRUE)
htmlwidgets::saveWidget(gg, "07_texas_persistent_hl.html", selfcontained = F, libdir = "libs")
```

<iframe src="07_texas_persistent_hl.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
class: inverse, center, middle

# Track data across multiple panels

---

### Highlight a specific year in 4 cities of interest: static version 

.pull-left[
```{r texas_tracking_static}
tx <- txhousing %>% filter(
  city %in% c("Austin", "Dallas", 
              "Houston", "Paris"))
p <- ggplot(tx, 
            aes(x = month, 
                y = median, 
                group = year)) + 
  geom_line(alpha = 0.2) +
  geom_line(data = filter(tx, year == 2010), #<<
            color = "red") +                 #<<
  facet_wrap(~city, ncol = 2) +
  theme(legend.position = "n")
plot(p)
```
]

.pull-right[
```{r ref.label="texas_tracking_static", fig.width=7, fig.height=7, fig.retina=3, echo = FALSE, eval=TRUE}
```
]

---
### Highlight a specific year in 4 cities of interest: dynamic version 

```{r texas_tracking_dynamic, echo = -(1:8)}
tx <- txhousing %>% 
  filter(city %in% c("Austin", "Dallas", "Houston", "Paris")) %>% 
  highlight_key(~year)
p <- ggplot(tx, aes(x = month, 
                    y = median, 
                    group = year)) + 
  geom_line() +
  facet_wrap(~city, ncol = 2)
ggplotly(p, tooltip = "year") %>% 
  highlight(on = "plotly_click", persistent = TRUE, 
            dynamic = TRUE, selectize = TRUE)
```

```{r save_texas_tracking_dynamic, include=FALSE, eval = generate_html}
tx <- txhousing %>% 
  filter(city %in% c("Austin", "Dallas", "Houston", "Paris")) %>% 
  highlight_key(~year)
p <- ggplot(tx, aes(x = month, 
                    y = median, 
                    group = year)) + 
  geom_line() +
  facet_wrap(~city, ncol = 2)
gg <- ggplotly(p, tooltip = "year", width = 800, height = 350) %>% 
  highlight(on = "plotly_click", selectize = TRUE, dynamic = TRUE, persistent = TRUE)
htmlwidgets::saveWidget(gg, "08_texas_tracking_dynamic.html", selfcontained = F, libdir = "libs")
```

<iframe src="08_texas_tracking_dynamic.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>
---

### Animation via the `frame` argument/aesthetic

```{r texas_animate}
p <- ggplot(tx, aes(x = month, y = median)) + 
  geom_line(aes(group = year)) +
  geom_line(aes(frame = year), color = "red") + #<<
  facet_wrap(~city, ncol = 2)
ggplotly(p)
```

```{r save_texas_animate, include=FALSE, eval = generate_html}
p <- ggplot(tx, aes(x = month, y = median)) + 
  geom_line(aes(group = year)) +
  geom_line(aes(frame = year), color = "red") + 
  facet_wrap(~city, ncol = 2) +
  theme(legend.position = "n")
gg <- ggplotly(p, width = 800, height = 350)
htmlwidgets::saveWidget(gg, "09_texas_animate.html", selfcontained = F, libdir = "libs")
```

<iframe src="09_texas_animate.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Gapminder

```{r gapminder_animate}
data(gapminder, package = "gapminder")
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) + #<<
  scale_x_log10()
ggplotly(p)
```

```{r save_gapminder_animate, include=FALSE, eval = generate_html}
data(gapminder, package = "gapminder")
p <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  labs(x = "GDP per capita", y = "Life expectancy") +
  scale_x_log10()
gg <- ggplotly(p, width = 800, height = 350, tooltip = "ids")
htmlwidgets::saveWidget(gg, "10_gapminder_animate.html", selfcontained = F, libdir = "libs")
```

<iframe src="10_gapminder_animate.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Gapminder (Hans Rosling's talk)

<div style="float:center ; width:80%">
<video width="854" height="480" controls>
  <source src="videos/hans_rosling.webm" type="video/webm">
</video>
</div> 

---
class: inverse, center, middle

# Saving a plotly object 

---
### Saving your interactive graphic

```{r dummy_plot, eval = TRUE}
p <- plot_ly(x = 1:10, y = 1:10) %>% add_markers()
```


- With `htmlwidgets::saveWidget`
```{r save_widget, echo = 1, eval = TRUE}
htmlwidgets::saveWidget(p, file = "my_widget.html")
cat("Size = ", round(file.size("my_widget.html")/1e6, 3), "MB")
unlink('my_widget.html')
```

The resulting html file is fully self-contained. 

- You can use `partial_bundle` to reduce the file size (by keeping only the parts of plolty.js used in the graphic)

```{r save_widget_partial, echo = 1, eval = TRUE}
htmlwidgets::saveWidget(partial_bundle(p), file = "my_widget.html")
cat("Size = ", round(file.size("my_widget.html")/1e6, 3), "MB")
unlink("my_widget.html")
```

---
### Saving multiple graphics

If you create multiple graphics (e.g. for a presentation), don't make every single one *self-contained* but specify a *shared lib directory*

```{r save_widget_not_self_contained, echo = 1:2, eval = TRUE}
htmlwidgets::saveWidget(p, file = "my_widget.html", 
                        selfcontained = FALSE, libdir = "my_libdir")
htmlwidgets::saveWidget(p, file = "my_other_widget.html", 
                        selfcontained = FALSE, libdir = "my_libdir")
size <- file.size("my_widget.html") + file.size("my_other_widget.html") + 3808e3
cat("Total size = ", round(size /1e6, 3), "MB")
unlink("my_widget.html")
unlink("my_other_widget.html")
unlink("my_libdir", recursive = TRUE, force = TRUE)
```

<i class="fa fa-exclamation-triangle"></i> If you share it with someone, be sure to share the lib folder <i class="fa fa-folder"></i> at the same time. 

---
class: middle, center, inverse

# Linking multiple views: highlight

---
### Highlight key 

`highlight_key` creates a *key* used to *link* data

```{r linking_example}
# declare the key and build a plotly object
tx <- txhousing %>% 
  highlight_key(~city) %>% 
  plot_ly(color = I('black')) %>% 
  group_by(city)

# initiate a plotly object (a view)
p1 <- tx %>% group_by(city) %>% add_lines(x = ~date, y = ~median)

# Initiate a second plotly object (a second view)
p2 <- tx %>% group_by(city) %>% 
  summarise(miss = sum(is.na(median))) %>%
  filter(miss > 0) %>% ungroup() %>% 
  mutate(city = forcats::fct_reorder(city, miss)) %>% 
  add_markers(x = ~miss, y = ~city)
```

---

```{r linking_example_left}
p1 <- tx %>% group_by(city) %>% 
  add_lines(x = ~date, y = ~median)
```

```{r save_linking_example_left, eval = generate_html, echo = FALSE}
tx <- txhousing %>% 
  highlight_key(~city) %>% 
  plot_ly(color = I('black')) %>% 
  group_by(city)
p1 <- tx %>% group_by(city) %>% 
  add_lines(x = ~date, y = ~median)
highlight(p1, width = 800, height = 350, 
          on = "plotly_click", dynamic = TRUE, selectize = TRUE) %>% 
  htmlwidgets::saveWidget(file = "11_linking_example_left.html", 
                          selfcontained = FALSE, libdir = "libs")
```

<iframe src="11_linking_example_left.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---

```{r linking_example_right}
p2 <- tx %>% group_by(city) %>% summarise(miss = sum(is.na(median))) %>%
  filter(miss > 0) %>% ungroup() %>% 
  mutate(city = forcats::fct_reorder(city, miss)) %>% 
  add_markers(x = ~miss, y = ~city)
```

```{r show_linking_example_right, eval = generate_html, echo = FALSE}
p2 <- tx %>% group_by(city) %>% summarise(miss = sum(is.na(median))) %>%
  filter(miss > 0) %>% ungroup() %>% 
  mutate(city = forcats::fct_reorder(city, miss)) %>% 
  add_markers(x = ~miss, y = ~city)
highlight(p2, width = 800, height = 350, 
          on = "plotly_click", dynamic = TRUE, selectize = TRUE) %>% 
  htmlwidgets::saveWidget(file = "12_linking_example_right.html", 
                          selfcontained = FALSE, libdir = "libs")
```

<iframe src="12_linking_example_right.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
### Combine both plots 

```{r subplot}
subplot(p2, p1, widths = c(0.2, 0.8), titleX = TRUE) %>%
  layout(showlegend = FALSE) %>%
  highlight(on = "plotly_click", dynamic = TRUE, selectize = TRUE)
```

```{r save_subplot, eval = generate_html, echo = FALSE}
subplot(p2, p1, widths = c(0.2, 0.8), titleX = TRUE) %>%
  layout(showlegend = FALSE, width = 800, height = 350) %>% 
  highlight(on = "plotly_click", dynamic = TRUE, selectize = TRUE) %>% 
  htmlwidgets::saveWidget(file = "13_subplot.html", selfcontained = FALSE, 
                          libdir = "libs")
```

<iframe src="13_subplot.html" width="100%" height="500" scrolling="no" seamless="seamless" frameBorder="0"> </iframe>

---
class: middle, center, inverse

# Linking multiple views: filters

---
### Filters

```{r filter_example}
library(crosstalk)
# No key specified, data are indexed by row number 
# Allows for the use of multiple keys
tx <- highlight_key(txhousing)
## Build plot
p <- ggplot(tx) + geom_line(aes(date, median, group = city))
gg <- ggplotly(p, dynamicTicks = TRUE)                     #<<
## Build filter
filter <- filter_select("id", "Select a city", tx, ~city)  #<<
## Combine filter and plot 
bscols(filter, gg, widths = c(12, 12))
```

---
### Filters

```{r ref.label="filter_example", echo = FALSE, cache = FALSE, eval = TRUE}
```

---
### Filters (advanced)

```{r filter_advanced_example}
tx <- highlight_key(txhousing)
## Many filters
widgets <- bscols(
  widths = c(12, 12, 12),
  filter_select("city", "Cities", tx, ~city),                #<<
  filter_slider("sales", "Sales", tx, ~sales),               #<<
  filter_checkbox("year", "Years", tx, ~year, inline = TRUE) #<<
)
## A plot 
p <- ggplot(tx, aes(x = date, y = median, group = city)) + 
  geom_line() + theme(legend.position = "none")
gg <- ggplotly(p)                                              #<<

## Combining both 
bscols(widgets, gg, widths = c(4, 8))
```

---
### Fiters (advanced)

```{r ref.label = "filter_advanced_example", echo = FALSE, cache = FALSE, eval = TRUE}
```

---
class: middle, center, inverse

# Plotly and other htmlwidgets

---
### Leaflet + Plotly

```{r leaflet_plotly}
library(leaflet)
## index data 
eqs <- highlight_key(quakes)

## Filter
stations <- filter_slider("station", "Number of Stations", eqs, ~stations)

## Plot 
p <- ggplot(eqs, aes(x = depth, y = mag)) +
  geom_point(alpha = 0.5)
gg <- ggplotly(p) %>% highlight(on = "plotly_selected")

## Map 
map <- leaflet(eqs) %>% 
  addTiles() %>% 
  addCircles()

## Combine
bscols(
  widths = c(6, 6, 3), 
  gg, map, stations
)
```

---
### Leaflet + Plotly

```{r show_leaflet_plotly, cache=FALSE, echo = FALSE, eval = TRUE}
eqs <- highlight_key(quakes)
stations <- filter_slider("station", "Number of Stations", eqs, ~stations)

p <- plot_ly(eqs, x = ~depth, y = ~mag) %>% 
  add_markers(alpha = 0.5) %>% 
  highlight("plotly_selected")

map <- leaflet(eqs) %>% 
  addTiles() %>% 
  addCircles()

bscols(
  widths = c(6, 6, 3), 
  p, map, stations
)
```

---
### DT + Plotly

```{r dt_plotly, cache = FALSE}
ir <- highlight_key(iris) 
bscols(
  ## Plot
  plot_ly(ir, x = ~Sepal.Width, y = ~Sepal.Length, color = ~Species) %>% 
    highlight(on = "plotly_selected", off = "plotly_deselect"), 
  ## datatable widget
  DT::datatable(ir)
)
```

---

```{r dt_plotly_show, echo = FALSE, eval=TRUE}
sketch = htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'ID'),
      th(rowspan = 2, 'Species'),
      th(colspan = 2, 'Sepal'),
      th(colspan = 2, 'Petal')
    ),
    tr(
      # lapply(rep(c('Length', 'Width'), 2), th) # for Paul/Michel
      th("Length"),
      th("Width"),
      th("Length"),
      th("Width")
    )
  )
))

ir <- iris %>% select(Species, Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) %>% 
  rename(sp = Species, spl = Sepal.Length, spw = Sepal.Width, pl = Petal.Length, pw = Petal.Width) %>% 
  highlight_key() 

p <- ggplot(ir, aes(x = spl, y = spw, color = sp)) + 
  geom_point() + 
  scale_color_discrete(name = "Species") + 
  labs(x = "Sepal Length", y = "Sepal Width") + 
  theme_minimal() + 
  theme(legend.position = "bottom")

bscols(
  ggplotly(p) %>% highlight(on = "plotly_select", off = "plotly_deselect"), 
  datatable(ir, extensions="Scroller", style="bootstrap", class="compact", width="100%",
            container = sketch, options=list(deferRender=TRUE, dom = "tp", pageLength = 10)), 
  widths = c(6, 6)
)
```

---

# Thanks!

Resources for more learning:

https://plotly-r.com/ <br />
https://rstudio.github.io/crosstalk/index.html <br />
 <br />
